A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-10 tex/codatadp.tex
More work on mu-values.

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Gc}{\mathsf{G}}
\newcommand{\Ac}{\mathsf{A}}
\newcommand{\Dc}{\mathsf{D}}
\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\quad \parallel \quad}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{[\![ #1 ]\!]}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{ [ #1 ] }
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}

@@

Recall that the $\mu$-notation can be used to represent (possibly cyclic) values taken from an initial algebra, \rem{elaborate?} 
e.g. $\mu x. \const{C}( \const{0}, x )$ represents the infinite sequence $\const{C}( \const{0}, \const{C}( \const{0}, \ldots ))$.
In this setting, a value can be viewed as a term-like structure where all subterms 
(which we will refer to as \emph{subvalues}) are either 
$\mu x. \const{C}( \vec v )$ for some constructor $\const{C}$ and subvalues $\vec v$
\footnote{We assume all such values are well-typed.}, or a variable $x$.
We will refer to values of this form as \emph{$\mu$-values}.
We say a $\mu$-value is \emph{closed} if it contains no free variables.
For convienience, we will represent the values of both datatypes and co-datatypes as closed $\mu$-values,
where the case of datatypes adds the restriction that none of its subvalues are variables.

To reason about the uniqueness of $\mu$-values, we introduce the following notions.
First, the \emph{interpretation of subvalue $u$ in $v$} is the value $v\interp{u}^\emptyset$ returned by the following,
where $x \mapsto \mu x. C( \vec t ) \in v$ indicates that $\mu x. C( \vec t )$ is the subvalue of $v$ that binds this occurrence of variable $x$.

\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
v\interp{x}^M & = & x & x \in M \\
v\interp{x}^M & = & v\interp{\mu x. C( \vec t )}^M & x \not\in M, x \mapsto \mu x. C( \vec t ) \in v  \\
v\interp{\mu x. C( \vec t )}^M & = & x & x \in M \\
v\interp{\mu x. C( \vec t )}^M & = & \mu x. C( v\interp{ \vec t }^{M \cup \{ x \}} ) & x \not\in M
\end{array}\]

We say a (closed) $\mu$-value $v$ of the form $\mu x. \const{C}( \vec w )$ is \emph{normal} if and only if 
$v\interp{u}^\emptyset$ is not equivalent to $v$ modulo renaming for all of its strict subvalues $u$,
and all of its strict closed subvalues are normal.
For instance, the value $v$ of the form $\mu x. \const{S}( \mu y. \const{S}( x ) )$ is not normal, 
since 
$v\interp{\mu y. \const{S}( x )}^\emptyset =$ 
$\mu y. \const{S}( v\interp{x}^{\{y\}} ) =$ 
$\mu y. \const{S}( v\interp{\mu x. \const{S}( \mu y. \const{S}( x ) ) }^{\{y\}} ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( v\interp{\mu y. \const{S}( x )}^{\{x,y\}} ) ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( y ) )$,
which is equivalent to $v$ modulo the renaming $\{ x \mapsto y, y \mapsto x \}$.

>>>>

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Gc}{\mathsf{G}}
\newcommand{\Ac}{\mathsf{A}}
\newcommand{\Dc}{\mathsf{D}}
\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\quad \parallel \quad}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{[\![ #1 ]\!]}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{ [ #1 ] }
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\nf}[1]{{#1}\!\!\downarrow}
\newcommand{\vsim}{\sim}

@@

Recall that the $\mu$-notation can be used to represent (possibly cyclic) values taken from an initial algebra, \rem{elaborate?} 
e.g. $\mu x. \const{C}( \const{0}, x )$ represents the infinite sequence $\const{C}( \const{0}, \const{C}( \const{0}, \ldots ))$.
In this setting, a value can be viewed as a term-like structure where all subterms 
(which we will refer to as \emph{subvalues}) are either 
$\mu x. \const{C}( \vec v )$ for some constructor $\const{C}$ and subvalues $\vec v$
\footnote{We assume all values are well-typed. 
Here, the type of $\mu x. \const{C}( \vec v )$ and $x$ is the return type of $\const{C}$.}, or a variable $x$.
We will refer to values of this form as \emph{$\mu$-values}.
We say a $\mu$-value is \emph{closed} if it contains no free variables.
For convienience, we will represent the values of both datatypes and co-datatypes as closed $\mu$-values,
where the case of datatypes adds the restriction that none of its subvalues are variables.

We introduce the following notions concerning $\mu$-values.
First, we write $v \vsim u$ if $\mu$-values $v$ and $u$ are equivalent modulo renaming of $\mu$-bound variables.
The \emph{interpretation of subvalue $u$ in $v$} is the value $v\interp{u}^\emptyset$ returned by the following recursive procedure,
where $x \mapsto \mu x. C( \vec t ) \in v$ indicates that $\mu x. C( \vec t )$ is the subvalue of $v$ that binds this occurrence of variable $x$.
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
v\interp{x}^M & = & v\interp{\mu x. C( \vec t )}^M & x \not\in M, x \mapsto \mu x. C( \vec t ) \in v  \\
v\interp{x}^M & = & x & \text{otherwise} \\
v\interp{\mu x. C( \vec t )}^M & = & \mu x. C( v\interp{ \vec t }^{M \cup \{ x \}} ) & x \not\in M \\
v\interp{\mu x. C( \vec t )}^M & = & x & \text{otherwise}
\end{array}\]
Given a $\mu$-value $v$ of the form $\mu x. C( \vec w )$, 
a subvalue $u$ of $v$ is a \emph{bisimilar subvalue} if and only if $v\interp{u}^\emptyset \sim v$.
We say $v$ is \emph{normal} if and only if all of its bisimilar subvalues are $x$,
and all of its strict subvalues that do not contain $v$ are also normal.
For instance, the value $v$ of the form $\mu x. \const{S}( \mu y. \const{S}( x ) )$ is not normal, 
since 
$v\interp{\mu y. \const{S}( x )}^\emptyset =$ 
$\mu y. \const{S}( v\interp{x}^{\{y\}} ) =$ 
$\mu y. \const{S}( v\interp{\mu x. \const{S}( \mu y. \const{S}( x ) ) }^{\{y\}} ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( v\interp{\mu y. \const{S}( x )}^{\{x,y\}} ) ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( y ) )$,
which is equivalent to $v$ modulo the renaming $\{ x \mapsto y, y \mapsto x \}$.
For any $\mu$-value $v$ of the form $\mu x. C( \vec w )$, we may recursively construct a normal value $\nf{v}$
(which we will call the \emph{normal form for $v$})
that is observationally equivalent to $v$.
We construct $\nf{v}$ by replacing all of the bisimilar subvalues of $v$ with $x$,
and by replacing all of the non-normal closed subvalues of $v$ with their corresponding normal forms.
For instance, $\nf{(\mu x. \const{S}( \mu y. \const{S}( x ) ))} = \mu x. \const{S}( x )$.
